### 2016-03-16 Manuscript/SW_early_life_telomeres.Rmd
Discussion and fogure edits

###Abstract
Understanding the short and long term costs of individual early-life experiences is fundamental to understanding life-history evolution. Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival. Thus, telomere dynamics may help us quantify individual variation in early-life costs, and enhance our understanding of how poor conditions in early life are related to later-life survival. We tested how telomere dynamics are related to spatiotemporal variation in early-life conditions and later-life survival in the Seychelles warbler (*Acrocephalus sechellensis*), across multiple cohorts spanning 14 years. we found that, in accordance with other studies, telomere length and loss are greatest in early life. We then show that juvenile telomere lengths varies markedly among cohorts, with average telomere length varying by . We found no evidence that early-life social environment (number of helpers, group size) or ecological conditions (territory quality) were related to telomere length, although we found tentative evidence that telomere length varied among summer and winter breeding seasons. Finally, we found that increased survival later in life was associated with longer telomeres in early life, but this effect was at the cohort, rather than individual, level. Our results highlight the inmportance of cohort effects in studies of telomere length.

@@

While it is clear that silver spoon effects can occur, we have little understanding of how and under what conditions early-life environments will affect adult fitness. However , we do expect the later-life consequences of a good or poor start to vary among individuals, populations and species because the phenotypic consequences of an adverse environment can depend on an individual's initial condition, and genetic or epigenetic makeup [@Hoffman2000; @Richards2006]. Moreover, it is not always possible to fully quantify what constitutes a good or bad environment, and any 'hidden' environmental variation may obscure relationships between the early-life conditions that are measured and the resulting adult phenotypes. Indeed, the pervasiveness of silver spoon effects varies between species [@Drummond2011], cohorts [@Reid2003] and sexes [@Wilkin2009]. We therefore need to understand, or at least be able to measure, how the environment affects individuals differentially within a population in order to elucidate the later-life consequences of early-life experience.

Telomeres - the protective caps on the ends of chromosomes - may provide a solution to this problem. Telomeres shorten with age [@Monaghan2006], and in response to oxidative stress, which can be elevated  due to environmental factors [@VonZglinicki2002]. When telomeres become critically short cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death  [@Wong2003]. This association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014]. While there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Barrett2011; @Simons2015], there is now excellent evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].

There is evidence from a range of taxa that the greatest rate of telomere loss occurs in early life [e.g. @Frenck1998; @Haussmann2003], and that the extent of this telomere shortening is influenced by the conditions experienced during that period [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015]. Importantly, early-life telomere dynamics have been associated with both short-term and late-life survival [@Heidinger2012; @Boonekamp2014], and with other parameters such as cognition [@Nettle2015a]. However, few studies have simultaneously analysed how telomere dynamics, early-life conditions and late-life survival are all related in a natural setting. Moreover, how early-life telomere dynamics vary over spatial and temporal scales is poorly understood.

@@

In this study we examine how telomeres link early-life environmental variation to late-life fitness in the Seychelles warbler. Because telomere dynamics are expected to reflect individual-level variation in the costs of early-life experiences, they may allow a more sensitive analysis of the effects of early-life environmental variation than would be possible with a direct comparison of how survival is affected by the early-life environment. With this in mind, we first conduct an exploratory analysis of how the environmental and social factors experienced in early life affect telomere dynamics. We then test the hypothesis that longer telomeres and lower rates of telomere shortening in early life are associated with greater survival.

@@

#####Molecular methods
For each sample, genomic DNA was extracted from a ~2 mm^2^ flake of preserved blood using the DNeasy Blood and Tissue Kit (Qiagen), following the manufacturer's protocol, with the modification of overnight lysis at 37^o^C and a final DNA elution volume of 80 $\mu$l.  Sex was determined using the PCR-based method outlined by Griffiths *et al.* [-@Griffiths1998]. Prior to telomere analysis, DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific), and the following thesholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 ratio has to be between 1.8 and 2 and, iii) the 260/230 ratio should be higher than 1.8. DNA integrity was further validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel, and all samples with evidence of DNA degradation were re-extracted or excluded.

All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement. We measured telomere length for all samples using a quantitative PCR (qPCR) assay of telomeres and a GPADH control gene, using the molecular methods outlined by Barrett *et al.* [-@Barrett2012], with one amendment. A change in batch of SYBR green forced us to raise the annealing temperature of the telomere reaction from 58^o^C to 61^o^C for the majority of samples. However, this did not affect final telomere length values (see below). We used the program LinRegPCR [@Ruijter2009] to correct for baseline fluorescence and calculate efficiencies and Cq values for each sample replicate. Averaging of technical repeats was carried out using custom-made R scripts (available as supplementary material), excluding samples with Cq values differing by >0.5. We then calculated relative telomere length (RTL) for each sample using equation 1 in Pfaffi *et al* [@Pfaffi2001]. We chose to use RTL rather than continuing with the previously used method for claculating absolute telomere length, as i) using RTL enabled us to run more samples per plate (as an oligo standard is not required), ii) the RTL method was less susceptible to batch effects (Appendix 1), and iii) very few other studies have calculated absolute telomere length, and our experience suggests that cross-species comparisons are unlikely to be reliable.

Inter-plate repeatability of final telomere lengths was assessed using the R package rptR [@Schielzeth2011], and for all subsequent analyses we used mean telomere length per sample where we had repeats.

Telomere lengths were measured using a total of `r nrow(juv)+nrow(Loss)` samples, . Of these, `r nrow(juv)` were taken cross-sectionally from birds caught within their first year of life, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`. For a subset of first-year birds (n = `r nrow(Loss)` individuals) we had longitudinal data, with an additional sample taken as an adult. For these individuals we calculated the rate of telomere loss by subtracting adult RTL from early-life RTL, and dividing this difference by the length of time (to the nearest 6 months) between sampling events.

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. Telomere length was log transfomred to fit assumptions of normality. We first explored how telomere length varied within the first year of life and over time using linear regression (telomere length vs age in months), and one-way ANOVA (telomere length vs cohort). With the longitudinal data, we tested how telomeres shorten with age by testing how telomere loss and rate of loss were related to the time interval between sampling events, using linear models. If telomeres are lost at a constant rate from early age, we expect the time between sampling events to be positively and linearly relaited to telomere loss no, and unrelated to rate of loss. If, however, telomeres are lost at a greater rate early in life we expect a non-linear or negative relationship between telomere loss and time between sampling events, and a significant (linear or non-linear) decrease in telomere rate of loss with time interval.

We then used a general linear mixed model approach to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012]. We first created a full model containing the following explanatory variables: age (in months), tarsus length, sex, territory quality, season (summer or winter) and the number of helping and non-helping subordinate birds present in the natal territory. We also included interaction terms between age and all the other variables as telomere dynamics are epxected to vary within the first year of life [@Heidinger2012]. As random effects we included cohort, territory ID and qPCR plate ID. Model selection was then performed and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. For individuals with longitudinal data (*n* = `r nrow(Loss)`) we repeated the above analyses of telomere dynamics, replacing telomere length with telomere loss as the response variable, and excluding the plate ID random effect (as each measurement of telomere loss was based on two or more measurements, and so run on multiple plates).

We used Cox regression, implemented in the survival package in R [@Therneau2011], to test whether survival was related to individual-level telomere length and rate of loss. For the cross-sectional data we used lifespan (in years) as survival time, while for the longitudinal data we used remaining lifespan from the second sampling event. We ran these models with cohort as a frailty term to exclude cohort effects.

###Results

#####Early-life telomere dynamics
Efficiecies (mean $\pm$) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively. Inter-plate repeatability of RTL, based on XX samples measured at least twice, was 0.78 (CI = 0.73-0.82).

RTL dereased with age within the first year of life (R^2^ = `r round2(summary(lm(RTL~Agemonths,data=juv))$r.squared)`; F = `r round2(anova(lm(RTL~Agemonths,data=juv))[1,4])`; P `r round3(anova(lm(RTL~Agemonths,data=juv))[1,5])`), and across the entire Seychelles warbler lifespan (R^2^ = `r round2(summary(lm(RTL~Agemonths,data=dd))$r.squared)`; F = `r round2(anova(lm(RTL~Agemonths,data=dd))[1,4])`; P `r round3(anova(lm(RTL~Agemonths,data=dd))[1,5])`; Fig. 1A). Longitudinal data showed that the rate of early-life telomere shortening also decreased with age (R^2^ = `r round2(summary(lm(TROC~Agemonths,data=Loss))$r.squared)`; F = `r round2(anova(lm(TROC~Agemonths,data=Loss))[1,4])`; P = `r round3(anova(lm(TROC~Agemonths,data=Loss))[1,5])`), suggesting that the highest rates of telomere loss occurred following the nestling stage (Fig. 1B).

Both early-life RTL and rate of loss varied significantly among cohorts (one-way ANOVA, telomere length: F = `r round2(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,4])`; P `r round3(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,5])`; Fig, 1C; telomere loss: F = `r round2(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,4])`; P `r round3(anova(lm(TROC~factor(FieldPeriodID),data = Loss))[1,5])`). Variation in median RTL over breeding seasons in fledglings and subadults was not related temporal variation in  territory quality (linear regression, R^2^ = `r round2(summary(lm(RTL~TQ,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(RTL~TQ,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(RTL~TQ,data = juvseason,na.action=na.exclude))[1,5])`), island-wide food availability (R^2^ = `r round2(summary(lm(RTL~Insect,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(RTL~Insect,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(RTL~Insect,data = juvseason,na.action=na.exclude))[1,5])`) or population density (R^2^ = `r round2(summary(lm(RTL~Density,data=juvseason,na.action=na.exclude))$r.squared)`; F = `r round2(anova(lm(RTL~Density,data = juvseason,na.action=na.exclude))[1,4])`; P = `r round2(anova(lm(RTL~Density,data = juvseason,na.action=na.exclude))[1,5])`).

#####Early-life environment and early-life telomere dynamics
The top model explaining variation in early-life RTL contained age, season (summer vs winter), tarsus length, and the interaction between age and tarsus length (Table S1). The top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+Season+Tarsus+Agemonths*Tarsus+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE)))`), and all three effects were statistically significant (Fig. 2A). RTL was higher in winter compared to summer seasons (Fig. 2B). Tarsus length was negatively related to RTL in nestlings, unrelated to RTL in fledglings, and positively related to RTL in subadults (Fig. 2C).

Model averaging results from the longitudinal data are shown in Figure 3A. The top model explaining telomere loss contained age and season (Table S2). This model was a significantly better fit than the null model ($\Delta$AICc = `r round2(AICc(lmer(TROC~1+(1|LayYear),data=Loss,REML = FALSE))-AICc(lmer(TROC~Agemonths+Season+(1|LayYear),data=Loss,REML = FALSE)))`). Juveniles born in winter seasons had higher rates of telomere shortening compared to birds born in summer seasons, but this effect was weak (Fig. 3B), although this effect was weak and confidence intervals overlapped zero (Fig. 3A). There was no evidence that tarsus length was related to telomere shortening in the same way that telomere length was (Fig. 3C).

@@

Telomere length in early life did not affect survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~RTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext~RTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,4])`), and there was no interaction effect between RTL and juvenile age on survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~RTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glm(SurvivedNext~RTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,4])`). To separate out cohort-level RTL affected survival we calculated cohort-level RTL (i.e. the mean for each cohort) and added this term to a logistic regression. However, we found no effect of RTL on survival to adulthood (estimate = `r round2(summary(glm(SurvivedNext~cohortTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext~cohortTL+Agemonths,data=juv,family = 'binomial'))$coefficients[2,4])`), and no interaction with juvenile age (estimate = `r round2(summary(glm(SurvivedNext~cohortTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glm(SurvivedNext~cohortTL*Agemonths,data=juv,family = 'binomial'))$coefficients[4,4])`).

Cox regression also showed no effect of individual-level or cohort-level RTL on survival later in life (individual-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL+Fledged)))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~RTL)))$coefficients[1,5])`;  cohort-level: estimate = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Fledged+frailty(LayYear))))$coefficients[1,1])`, P = `r round2(summary(coxph(with(juv9,Surv(Lifespan,Died)~cohortTL+Fledged+frailty(LayYear))))$coefficients[1,6])`). Using the longitudinal data, we found that the amount of telomere shortening experienced in early life had no effect on survival (estimate `r round3(summary(coxph(with(subset(Loss,LayYear<2009),Surv(RemainingLife2+1,Died)~TROC+Agemonths)))$coefficients[1,1])`, P = `r round3(summary(coxph(with(subset(Loss,LayYear<2009),Surv(RemainingLife2+1,Died)~TROC+Agemonths)))$coefficients[1,5])`).

###Discussion
Here we use the long-term study of a closed population of Seychelles warblers to assess the relationships between early-life conditions, telomere length and survival. We find RTL is subject to strong cohort effects

>>>>

###Abstract
Understanding the short and long term costs of individual early-life experiences is fundamental to understanding life-history evolution. Telomeres, the protective caps at the ends of chromosomes, shorten in response to oxidative stress, and telomere shortening is correlated with reduced survival. Thus, it is possible that telomere dynamics may help us quantify individual variation in early-life costs, and enhance our understanding of how poor conditions in early life are related to later-life survival. We tested how telomere dynamics are related to spatio-temporal variation in early-life conditions and later-life survival in the Seychelles warbler (*Acrocephalus sechellensis*), using a large dataset of `r nrow(juv)` juveniles from multiple cohorts spanning `r max(juv$LayYear)-min(juv$LayYear)` years. Despite cross-sectional data showing that telomere length decreases with age, and most that telomere length decreases most rapidly early in life, using longitudinal data we found that significant reductions in telomere length can only be detected within the first few months of life - after this, longitudinal changes in telomere length were not significantly different from zero. Juvenile telomere length varied markedly among cohorts, and this cohort variation was related to the season (birds born in winter seasons had longer telomeres), and to population density (birds bird in years with higher populaiton density had longer telomered). Within cohorts, we found no evidence that the early-life social environment or ecological conditions were related to telomere length or rates of shortening. Finally, we found no relationship between early-life and survival to adulthood. Our data suggest that telomeres are subject to strong population-level variation, but suggest that the resolution of telomeres as a useful biomarker of individual-level cost is likely to be age-dependent, and in many cases limited.

@@

While it is clear that silver spoon effects can occur, we have little understanding of how and under what conditions early-life environments will affect adult fitness. We expect the later-life consequences of a good or poor start to vary among individuals, populations and species because the phenotypic consequences of an adverse environment can depend on an individual's initial condition, and genetic or epigenetic makeup [@Hoffman2000; @Richards2006]. Moreover, it is not always possible to fully quantify what constitutes a good or bad environment, and any 'hidden' environmental variation may obscure relationships between the early-life conditions that are measured and the resulting adult phenotypes. Indeed, the pervasiveness of silver spoon effects varies between species [@Drummond2011], cohorts [@Reid2003] and sexes [@Wilkin2009]. We therefore need to understand, or at least be able to measure, how the environment affects individuals differentially within a population in order to elucidate the later-life consequences of early-life experience.

Telomeres - the protective caps on the ends of chromosomes - may provide a solution to this problem. Telomeres shorten with age [@Monaghan2006], and in response to oxidative stress, which can be elevated  due to environmental factors [@VonZglinicki2002]. When telomeres become critically short cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death  [@Wong2003]. This association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology [reviewed in @Horn2010; @Haussmann2010; @Monaghan2014]. While there is little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015], there is mounting evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [@Monaghan2014; @Schultner2014; @Asghar2015].

There is evidence from a range of taxa that the greatest rate of telomere shortening occurs in early life [e.g. @Frenck1998; @Haussmann2003], and that the extent of this shortening is influenced by the conditions experienced during that period [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015]. Early-life telomere dynamics have been associated with both short-term and late-life survival [@Heidinger2012; @Boonekamp2014], and with other parameters such as cognition [@Nettle2015a]. However, few studies have simultaneously analysed how telomere dynamics, early-life conditions and late-life survival are all related in a natural setting. Moreover, how early-life telomere dynamics vary over spatial and temporal scales is poorly understood.

@@

In this study we examine how telomeres are related to early-life environmental variation and short and long-term survival in the Seychelles warbler. Because telomere dynamics are expected to reflect individual-level variation in the costs of early-life experiences, they may allow a more sensitive analysis of the effects of early-life environmental variation than would be possible with a direct comparison of how survival is affected by the early-life environment. With this in mind, we first conduct an exploratory analysis of how the environmental and social factors experienced in early life affect telomere dynamics. We then test the hypothesis that longer telomeres and lower rates of telomere shortening in early life are associated with greater survival.

@@

#####Molecular methods
For each sample, genomic DNA was extracted from a ~2 mm^2^ flake of preserved blood using the DNeasy Blood and Tissue Kit (Qiagen), following the manufacturer's protocol, with the modification of overnight lysis at 37^o^C and a final DNA elution volume of 80 $\mu$l.  Sex was determined using the PCR-based method outlined by Griffiths *et al.* [-@Griffiths1998]. Prior to telomere analysis, DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific), and the following thresholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 ratio has to be between 1.8 and 2 and, iii) the 260/230 ratio should be higher than 1.8. DNA integrity was further validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel, and all samples with evidence of DNA degradation were re-extracted or excluded.

All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement. We measured telomere length for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, using the molecular methods outlined by Barrett *et al.* [-@Barrett2012], with one amendment. A change in batch of SYBR green forced us to raise the annealing temperature of the telomere reaction from 58^o^C to 61^o^C for the majority of samples. However, this did not affect final telomere length values (see below). We used the program LinRegPCR [@Ruijter2009] to correct for baseline fluorescence and calculate efficiencies and Cq values for each sample replicate. Averaging of technical repeats was carried out using custom-made R scripts (available as supplementary material), excluding samples with Cq values differing by >0.5. We then calculated relative telomere length (RTL) for each sample using equation 1 in Pfaffi *et al* [@Pfaffi2001]. We chose to use RTL rather than continuing with the previously used method for calculating absolute telomere length [@Barrett2012], as i) using RTL enabled us to run more samples per plate (as an oligo standard is not required), ii) the RTL method was less susceptible to batch effects (Appendix 1), and iii) very few other studies have calculated absolute telomere length, and our experience suggests that cross-species comparisons are unlikely to be reliable. Inter-plate repeatability of final telomere lengths was assessed using the R package rptR [@Schielzeth2011], and for all subsequent analyses we used mean telomere length per sample where we had repeats.

For a subset of first-year birds for which we had longitudinal data, with an additional sample taken as an adult. For these individuals we calculated the within-individual change in RTL by subtracting adult RTL from early-life RTL (hereafter $\Delta$RTL). Negative values of $\Delta$RTL reflect telomere shortening, while positive values reflect telomere lengthening.

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. Because we are using a different measure of telomere length to the previous study of adult telomere dynamics in this species [@Barrett2013], as well as including many additional samples, we first re-tested, using all available samples, whether telomere length decreases with age and $\Delta$age [a longitudinal measure based on within-subsject centring; @VandePol2009] in the Seychelles warbler, using linear mixed models, with individual ID as a random effect. For the analysis of $\Delta$age we we only included birds who had been sampled at least twice. We then explored how telomere length varied within the first year of life (telomere length vs age in months) and over time (telomere length vs cohort ID). For this analysis we had bvery few birds with multiple samples, so we used linear models including only the first sample for each individual.

To assess our resolution for detecting telomere shortening, we tested, for each age class, whether $\Delta$RTL was significantly lower than zero, using one-sample t-tests. We then tested how $\Delta$RTL was related to the time interval between sampling events, using linear models. A significantly negative slope indicates that telomere shortening occurs linearly over time, while the time point at which $\Delta$RTL is significantly lower than zero can be considered the time period over which we have the resolution to detect reductions in telomere length.

We used general linear mixed models with model averaging to explore how spatial variation in early-life environmental and social conditions influenced telomere length within cohorts. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012]. We first created a full model containing the following explanatory variables: age (in months), tarsus length, sex, territory quality, season (summer or winter) and the number of helping and non-helping subordinate birds present in the natal territory. We also included interaction terms between age and all the other variables as telomere dynamics are expected to vary within the first year of life [@Heidinger2012]. As random effects we included cohort, territory ID and qPCR plate ID. Model selection was then performed and a top model set defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model-averaged coefficients, confidence intervals and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. For individuals with longitudinal data (*n* = `r nrow(Loss)`) we repeated the above analyses of telomere dynamics, replacing telomere length with telomere loss as the response variable, and excluding the plate ID random effect (as each measurement of telomere loss was based on two or more measurements, and so run on multiple plates).

We used generalised linear mixed models to test whether telomere length/loss were related to survival to adulthood. Models were set up with a binomial error structure, and included RTL/$\Delta$RTL and juvenile age as explanatory variable, along with the same random effects as for the model averaging (see above). We then used Cox regression, implemented in the survival package in R [@Therneau2011], to test whether long-term survival was related to individual-level telomere dynamics. For the cross-sectional data we used lifespan (in years) as survival time, while for the longitudinal data we used remaining lifespan from the second sampling event. We ran these models with cohort as a frailty term to exclude cohort effects, and only included cohorts in which at least 70 percent of individuals had died (all years up to and including 2007).

###Results
We measured telomere lengths using a total of `r nrow(dd)` unique samples from juvenile and adult Seychelles warblers. Of these, `r nrow(juv)` samples were taken cross-sectionally from birds caught within their first year of life, and of these juvenile birds, we had additional longitudinal samples for `r nrow(Loss)`. Efficiencies (mean $\pm$) for our telomere and GAPDH reactions were (`r round2(mean(dd$EffTelomere))` $\pm$ `r round2(sd(dd$EffTelomere))`) and (`r round2(mean(dd$EffGAPDH))` $\pm$ `r round2(sd(dd$EffGAPDH))`) respectively. Inter-plate repeatability of RTL, based on 392 samples measured at least twice, was 0.78 (CI = 0.73-0.82).

#####Early-life RTL
Considering the entire Seychelles warbler lifespan, RTL decreased with age (t = `r round2(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~Agemonths,data=dd,random=~1|BirdID))$tTable[2,5])`; Fig. 1A), and with $\Delta$age (t = `r round2(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,4])`; P `r round3(summary(lme(RTL~DeltaAge,data=subset(dd,DeltaAge!=0),random=~1|BirdID))$tTable[2,5])`; Fig. 1B). RTL also decreased with age within the first year of life (i.e. excluding adult samples; R^2^ = `r round2(summary(lm(RTL~Agemonths,data=juv))$r.squared)`; t = `r round2(summary(lm(RTL~Agemonths,data=juv))$coefficients[2,3])`; P `r round3(summary(lm(RTL~Agemonths,data=juv))$coefficients[2,4])`). Early-life RTL varied significantly among cohorts (one-way ANOVA, telomere length: F = `r round2(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,4])`; P `r round3(anova(lm(RTL~factor(FieldPeriodID),data = juv))[1,5])`), and cohort-level variation in juvenile RTL (taken as the median for each cohort) was positively related to adult population density (R^2^ = `r round2(summary(lm(RTL~Density,data=juvseason,na.action=na.exclude))$r.squared)`; t = `r round2(summary(lm(RTL~Density,data = juvseason,na.action=na.exclude))$coefficients[2,3])`; P = `r round2(anova(lm(RTL~Density,data = juvseason,na.action=na.exclude))[1,5])`; Fig. 1C). Variation in cohort-level RTL was not related temporal variation in  territory quality (linear regression, R^2^ = `r round2(summary(lm(RTL~TQ,data=juvseason,na.action=na.exclude))$r.squared)`; t = `r round2(summary(lm(RTL~TQ,data = juvseason,na.action=na.exclude))$coefficients[2,3])`; P = `r round2(anova(lm(RTL~TQ,data = juvseason,na.action=na.exclude))[1,5])`), or island-wide food availability (R^2^ = `r round2(summary(lm(RTL~Insect,data=juvseason,na.action=na.exclude))$r.squared)`; t = `r round2(summary(lm(RTL~Insect,data = juvseason,na.action=na.exclude))$coefficients[2,3])`; P = `r round2(anova(lm(RTL~Insect,data = juvseason,na.action=na.exclude))[1,5])`).

The top model explaining variation in early-life RTL contained age, sex, season (summer vs winter), tarsus length, and the interaction between age and tarsus length (Table S1). The top model was much better supported than a null model ($\Delta$AICc = `r round2(AICc(lmer(RTL~1+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE))-AICc(lmer(RTL~Agemonths+Sex+Season+Tarsus+Agemonths*Tarsus+(1|LayYear)+(1|TerritoryID),data=juv,REML = FALSE)))`), and with the exception of the sex and tarsus effects, model-averaged confidence intervals did not overlap zero (Fig. 2A). RTL was higher in winter compared to summer seasons (Fig. 2B). Tarsus length was negatively related to RTL in nestlings, unrelated to RTL in fledglings, and positively related to RTL in subadults (Fig. 2C). Males had slightly longer telomeres than females, prarticularly at the fledgling and subadult stages (Fig. 2D), although this effect was weak.

#####Early-life telomere shortening
Longitudinal data showed that early-life RTL was positively, but weakly, related to adult RTL (R^2^ = `r round3(summary(lm(RTL1~RTL+Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(RTL1~RTL+Agemonths,data=Loss))$coefficients[2,3])`; P = `r round3(summary(lm(RTL1~RTL+Agemonths,data = Loss))$coefficients[2,4])`; Fig. 3A). $\Delta$RTL was significantly lower than zero in nestlings (t = `r round2(t.test(subset(Loss,Agemonths == 1)$Loss)$statistic)`; P `r round3(t.test(subset(Loss,Agemonths == 1)$Loss)$p.value)`), but not in fledglings (t = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 6)$Loss)$p.value)`) or subadults (t = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$statistic)`; P = `r round2(t.test(subset(Loss,Agemonths == 10)$Loss)$p.value)`). In line with this, $\Delta$RTL increased with age in early life (R^2^ = `r round2(summary(lm(Loss~Agemonths,data=Loss))$r.squared)`; t = `r round2(summary(lm(Loss~Agemonths,data=Loss))$coefficients[2,3])`; P `r round3(summary(lm(Loss~Agemonths,data = Loss))$coefficients[2,4])`), suggesting that telomere shortening was highest shortly after the nestling phase. In nestlings, $\Delta$RTL  did not vary with the time between sampling events (R^2^ `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$r.squared)`; t = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$coefficients[2,3])`; P = `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths ==1)))$coefficients[2,4])`). However, for fledglings and subadults $\Delta$RTL decreased with the length of time between sampling events (R^2^ = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$r.squared)`; t = `r round2(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$coefficients[2,3])`; P = `r round3(summary(lm(Loss~TimeDiff,data=subset(Loss,Agemonths !=1)))$coefficients[2,4])`). In flesdglings and subadults, $\Delta$RTL did not decrease below zero until ~3 years between sampling events was allowed, and confidence limits overlapped zero regardless of the amount of time between sampling events (Fig 3B). Together, these results suggest that the majority of observable telomere shortening in Seychelles warblers occurs in the first six months of life. We therefore restricted the remaining analyses of $\Delta$RTL to birds sampled as nestlings.

$\Delta$RTL did not differ significantly among cohorts (F = `r round2(anova(lm(Loss~factor(LayYear),data = subset(Loss,Agemonths == 1)))[1,4])`; P `r round3(anova(lm(Loss~factor(LayYear),data = subset(Loss,Agemonths == 1)))[1,5])`), although power for this analysis was limited as longitudinal sample sizes within cohorts was low (Fig. 3C). Model selection revealed that the top model explaining $\Delta$RTL was the null model (Table S2), and confidence limits for all explanatory terms overlapped zero (Fig. 4A).

@@

Early-life RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL+Agemonths+Tarsus+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction effect between RTL and juvenile age on survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~RTL*Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[4,4])`). To test whether cohort-level effects on RTL influenced survival we calculated cohort-level RTL (i.e. the median RTL for each cohort) and added this term to a GLMM, and found that cofort-level RTL was not related to survival to adulthood (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL+Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL+Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[2,4])`), with no interaction with juvenile age (estimate = `r round2(summary(glmer(SurvivedNext~cohortTL*Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[4,1])`; P = `r round2(summary(glmer(SurvivedNext~cohortTL*Agemonths+(1|TerritoryID)+(1|LayYear),data=juv13,family = 'binomial'))$coefficients[4,4])`). Using the longitudinal data, we found that $\Delta$RTL had no effect on survival to the year after the second sampling event (estimate = `r round2(summary(glm(SurvivedNext2~Loss,data=subset(subset(Loss,Agemonths == 1),LayYear<2013),family = 'binomial'))$coefficients[2,1])`; P = `r round2(summary(glm(SurvivedNext2~RTL+Agemonths,data=subset(subset(Loss,Agemonths == 1),LayYear<2013),family = 'binomial'))$coefficients[2,4])`).
